Knowing your role: Varying expectations in software organizations

### 2011-09-27 soulgem.tex
Third draft, Intro.

\begin{document}
\title{Working Title: Roles and Expectations in Software Organizations}

@@

Poeple divide their labour in order to produce software systems. Some write production code, others test it, and still others try to ensure that the product satisfies a real need. Researchers and practitioners usually abstract these responsibilities into a number of role labels (such as ``developer,'' ``tester,'' and ``product manager''); we use these abstractions to refer to the set of activities that we expect them to carry out. These labels are common everywhere in our academic literature, as well as in textbooks, books for practitioners, human resources postings, blog posts, and everyday conversation.

Despite this commonplace use, we have no clear indication that the people using these labels refer to the same meanings. For instance, what does a ``developer'' actually need to do in order to fulfill her responsibilities? What should we expect her to be unable to do? And would the answer be generalizable throughout the many areas in which the software industry operates?

In some domains, especially the more mature or critical ones, the division of labour is fairly formal and clear. Surgeons, anesthesiologists, nurses, and patients generally know what to expect from each other, in and out of the operating room. Knowing that someone is a nurse will bring out a set of expectations of behaviour regarding that person that have a good probability of being shared across the board. But although in our own community we assign labels just as directly and unthinkingly as in the medical domain, we do not realize (or such is the premise of our paper) that we may be talking about very different things, based on our varying assumptions, experiences, and cultures.

This ontological drift \cite{Robinson1991} in our conceptualization of software development roles has not been studied in depth so far, and it has unexplored repercussions for researchers and practitioners alike. Poorly defined roles may lead to dysfunction in a project team, and thus negatively impact the quality of the software. Researchers may prescribe practices to be carried out by one role without realizing that the role is usually shared and spread out among a group of people. Practitioners may not know how a newcomer's experience as a developer in a different organization will map into their own, or how to train him \cite{Dagenais2010}. Their communication tools may fail to be adopted if they satisfy the needs of some roles but not of others. For an abstraction as heavily used as role definitions in software organization, a closer look should prove beneficial.

In this paper we do not attempt to provide a rigid definition of roles in software organizations; in fact we argue that such a task has poor prospects of success. Instead, (a) we examine the concepts of roles, positions, and expectations in software development groups, providing a theoretical framework based on sociological and organizational research, (b) we report the findings from a case study at two software organizations in which we used this theoretical framework to ground our observations and interviews, and (c) we discuss the implications for research and practice from the framework and from our findings.

>>>>

\begin{document}
\title{Knowing your role: Varying expectations in software organizations}

@@

People divide their labour in order to produce software systems. Some write code, others test it, and still others ensure that the product satisfies a real need. Researchers and practitioners abstract these responsibilities into a number of role labels (such as ``developer,'' ``tester,'' and ``product manager''); we use these abstractions to refer to the set of activities that we expect them to carry out. These labels are common in our academic literature, as well as in textbooks, books for practitioners, human resources postings, blog posts, and everyday conversation.

Despite their widespread use, we have no clear indication that the people using these labels refer to the same meanings. For instance, what does a ``developer'' actually do in order to fulfill her responsibilities? What bounds her activity? And would the answer be generalizable throughout the many areas in which the software industry operates?

In some domains, especially the more mature or critical ones, the division of labour is fairly formal and clear. Surgeons, anesthesiologists, and nurses generally know what to expect from each other, in and out of the operating room. Knowing that someone is a nurse will bring out a set of expectations of behaviour regarding that person that have a good probability of being shared across the board. But although in our community we assign labels just as effortlessly as in the medical domain, we do not realize (or such is the premise of our paper) that we may be talking about different things, based on our varying assumptions, experiences, and cultures.

This ontological drift \cite{Robinson1991} in our conceptualization of software development roles has not been studied in depth so far, and it has unexplored repercussions for researchers and practitioners alike. Poorly defined roles may lead to dysfunction in a project team, and thus negatively impact the quality of the software. Researchers may prescribe practices to be carried out by one role without realizing that the role is usually shared and spread out among a group of people. Practitioners may not know how a newcomer's experience as a developer in a different organization will map into their own, or how to train him \cite{Dagenais2010}. Their communication tools may fail to be adopted if they satisfy the needs of some roles but not of others. For an abstraction as heavily used as roles in software organization, a closer look should prove beneficial.

In this paper we do not provide a rigid definition of roles in software organizations; in fact we argue that such a task has poor prospects of success. Instead, (a) we examine the concepts of roles, positions, and expectations in software groups, providing a theoretical framework based on interdisciplinary research, (b) we report the findings from a case study at two software organizations in which we used this framework to ground our observations and interviews, and (c) we discuss the implications for research and practice from the framework and from our findings.
